Compare NE-combined vs. NE-separated reference

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
notebook cleanups at render time

### UMAP

@@

First, we'll show all cell types:
```{r}
# remove NAs before plotting
ggplot(tidyr::drop_na(singler_long_df)) +
  aes(x = UMAP1, y = UMAP2, color = singler_label) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_manual(values = nbatlas_colors) +
  facet_wrap(vars(model)) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(size = 1, alpha = 1)))
```

Next, we'll show this faceted only for the `NE` cell types
```{r}
singler_ne_df <- singler_long_df |>
  dplyr::filter(stringr::str_detect(singler_label, "Neuroendocrine"))

@@

The follow table shows relationship between model annotations:

```{r}
table(
  singler_df$singler_label_combined,
  singler_df$singler_label_separated
)
```

@@

```{r}
# Define the marker genes
ne_marker_genes <- model_separated$markers$full$Neuroendocrine |>
  purrr::reduce(c) |>
  unique()

ne.tumor_marker_genes <- model_separated$markers$full$`Neuroendocrine-tumor` |>
  purrr::reduce(c) |>
  unique()

@@

```{r, fig.width = 14, message=F, warning=F}
# plot the NE-only genes
plot1 <- scater::plotExpression(
  sce[ne_only, neuro_barcodes],
  features = ne_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5,
  ncol = 6 # facet cols
) +
  stat_summary(shape = 18) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine marker genes")

# plot the NE-tumor--only genes
plot2 <- scater::plotExpression(
  sce[ne.tumor_only, neuro_barcodes],
  features = ne.tumor_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5,
  ncol = 6 # facet cols
) +
  stat_summary(shape = 18) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine-tumor marker genes")

@@

How many of the `NE` markers in the combined model (again, out of 10 total per cell type comparison!) are _also included_ in the separated model's `NE` (or `NE-tumor`) marker gene lists?

@@

combined_markers |>
  # only comparisons to other groups of cells
  purrr::discard_at("Neuroendocrine") |>
  purrr::imap(
    \(combined_marker_genes, cell_type) {
      separated_model_markers <- c(
        # upregulated in NE compared to cell_type
        ne_markers[[cell_type]],
        # upregulated in NE-tumor compared to cell-type
        ne.tumor_markers[[cell_type]]
      ) |> unique()

      # how many combined are in that separated list?
      sum(combined_marker_genes %in% separated_model_markers)
    }
  ) |>
  as.data.frame() |>
  t()
```

>>>>

### UMAPs

@@

First, we'll show a UMAP of all cell types for both the combined and separated models:

```{r}
# remove NAs before plotting
ggplot(tidyr::drop_na(singler_long_df)) +
  aes(x = UMAP1, y = UMAP2, color = singler_label) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_manual(values = nbatlas_colors) +
  facet_wrap(vars(model)) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(size = 1, alpha = 1)))
```

Next, we'll show this UMAP in a faceted view for _only_ the `NE` cell types.
This will help us see to what extent `NE` and `NE-tumor` cells are overlapping in the separated model, and how they are laid out compared to the combined model.

```{r}
singler_ne_df <- singler_long_df |>
  dplyr::filter(stringr::str_detect(singler_label, "Neuroendocrine"))

@@

The follow table shows relationship between model annotations.
The separated model is shown in rows, and the combined model in columns.
Each value is the count of cells with the given combination of labels.

```{r}
table(
  singler_df$singler_label_separated,
  singler_df$singler_label_combined
)
```

@@

```{r}
# Define the marker genes
ne_marker_genes <- ne_markers |>
  purrr::reduce(c) |>
  unique()

ne.tumor_marker_genes <- ne.tumor_markers |>
  purrr::reduce(c) |>
  unique()

@@

```{r, fig.width = 12, fig.height = 14, message=F, warning=F}
# plot the NE-only genes
plot1 <- scater::plotExpression(
  sce[ne_only, neuro_barcodes],
  features = ne_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5,
  ncol = 6 # facet cols
) +
  stat_summary(shape = 18) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  ggtitle("Neuroendocrine marker genes")

# plot the NE-tumor--only genes
plot2 <- scater::plotExpression(
  sce[ne.tumor_only, neuro_barcodes],
  features = ne.tumor_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5,
  ncol = 6 # facet cols
) +
  stat_summary(shape = 18) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  ggtitle("Neuroendocrine-tumor marker genes")

@@

How many of the `NE` markers in the combined model (again, out of 10 total per cell type comparison!) are _also included_ in the separated model's `NE` or `NE-tumor` marker gene lists?
The table below shows the counts for how many of the combined markers appear in the given separated model's marker gene lists.

@@

combined_markers |>
  # only comparisons to other groups of cells
  purrr::discard_at("Neuroendocrine") |>
  purrr::imap(
    \(combined_marker_genes, cell_type) {
      data.frame(
        n_markers_in_NE = sum(combined_marker_genes %in% ne_markers[[cell_type]]),
        n_markers_in_NE_tumor = sum(combined_marker_genes %in% ne.tumor_markers[[cell_type]])
      )
    }
  ) |>
  purrr::list_rbind(names_to = "cell_type")
```
